Highly Available Transactions: Virtues and Limitations

### 2013-03-30 writeup/hat-analysis.tex

\begin{quote}
``...one should not throw out the C so quickly, since there are real
  error scenarios where CAP does not apply and it seems like a bad
  tradeoff in many of the other situations.''---Michael
  Stonebraker~\cite{stonebraker2010errors}
\end{quote}

HAT systems---providing transactions with transactional availability
with high availability ($1$-available) or sticky high availability
($1$-sticky-available)---offer substantial latency and availability
benefits, yet they come with a cost to achievable semantics. In this
section, we delineate which of several ACID, distributed consistency,
and session consistency levels can be achieved with high availability
(e.g., transactional atomicity, variants of Repeatable Read isolation,
and many session guarantees) and sticky high availability (e.g.,
read-your-writes, PRAM, and causal consistency) and which cannot
(e.g., preventing Lost Update and Write Skew).  We present a full
summary of these results in Section~\ref{sec:hat-summary}.

@@

To begin, Adya's \textit{Read Uncommitted} isolation requires that
transactions are totally ordered and that writes within transactions
are ordered consistently with this order (prohibiting ``Dirty
Writes,'' or $G0$)~\cite{adya}. If two transactions write to the same
set of data items, then the final database state cannot contain the
``earlier'' transaction's writes to the data items. For example, in
the below example, $T_3$ should eventually only read $a=b=1$ or
$a=b=2$ but not $a=2, b=1$ or $a=1, b=2$:
\begin{align*}
\small
T_1 &: w_x(1)~w_y(1)
\\T_2 &: w_x(2)~w_y(2)
\\T_3 &: r_x(a)~r_y(b)\\[-2em]
\end{align*}
We will strengthen this property in later models, which will prevent,
for instance, reading $a=2, b=1$ at any time (not just
``eventually''). The difference between this total order on
transactions and the total order required by serializability is that
the Read Uncommitted total order is completely arbitrary (i.e., an
order must exist), whereas a serializable total order must be
equivalent to a serial execution. It is easily achieved via applying
the same logical timestamp to each update in a transaction and
applying a ``last writer wins'' conflict reconciliation policy at each
replica.

\textit{Read Committed} isolation requires that transactions do not
read uncommitted versions of data items (prohibiting both ``Dirty
Writes''---as above---and ``Dirty Reads'' phenomena; captured by
Adya's $G1a-c$, ANSI's $P1$, and ``broad'' $P1$ (2.2) from Berenson et
al.). For instance, in the example below, $T_3$ should never see
$a=1$, and, if $T_2$ aborts, $T_3$ should never read $a=3$:
\vspace{-.5em}
\begin{align*}
\small
T_1 &: w_x(1)~w_x(2)
\\T_2 &: w_x(3)\\
T_3 &: r_x(a)\\[-2em]
\end{align*}
It is fairly easy to prvent ``Dirty Reads'': if each transaction never
writes uncommitted data to the database, then transactions will never
read each others' dirty data. Clients can buffer their writes until
they commit, or, alternatively, can send them to servers, who will not
serve new writes until clients notify them that the writes have been
committed.

\textit{Repeatable Read} isolation is a contentious property. As
Berenson et al. discuss~\cite{ansicritique}, Gray's original
Repeatable Read lock-based implementation provides substantially
richer guarantees than those that are required by the ANSI SQL
specification~\cite{gray-isolation}. Gray~\cite{gray-isolation},
Berenson et al.~\cite{ansicritique}, and Adya~\cite{adya} all
interpret Repeatable Read as providing serializability for all
operations except predicate-based reads. However, in its description
of Repeatable Read, the ANSI SQL specification provides a useful
property which, although is not true to Gray's spirit of Repeatable
Read, is also commonly found in distributed consistency models: the
ANSI Repeatable Read guarantee requires that, along with observing
Read Committed isolation, each transaction will only read one version
of each data item that it did not itself produce (preventing ``Fuzzy
Read,'' or P2). In the example below, $T_3$ must read $a=1$:
\begin{align*}
\small
T_1 &: w_x(1)
\\T_2 &: w_x(2)
\\T_3 &: r_x(1)~r_x(a)
\end{align*}
This isolation property is a literal interpretation of the phrase
``Repeatable Read'': unless a transaction modifies a given data item,
the observable value of the data item should not change during the
transaction. By itself, this property is rather weak---we can always
read $\bot$ for each data item. However, when coupled with additional
properties, like transactional atomicity, it is stronger: the repeated
reads must obey additional ordering constraints. In fact, distributed
systems often describe a ``consistent snapshot'' across a set of
related events as a \textit{cut} across a set of participants or, in
our case, data items. Accordingly, to capture the notion that the
transaction should read from a (non-changing) consistent cut over data
items (and to disambiguate from the aforementioned stronger Repeatable
Read properties), we call ANSI SQL ``Repeatable Read'' (i.e.,
preventing P2) \textbf{Cut Isolation}. It is possible to satisfy cut
isolation with high availability by caching appropriate versions for
each of a transaction's reads or, alternatively, using
multi-versioning on each server and ensuring that each of a
transaction's successive reads return the same version of each data
item.

We can further consider two variants of Cut Isolation. The first is
relatively straightforward and provides a cut across individual data
items. We will call this \textit{Item Cut Isolation}. The second
provides a cut over individual data items plus logical ranges of data
items, or predicate-based reads. This prevents the \textit{Phantom
  Problem}, whereby two successive predicate-based reads return
different data~\cite{gray-isolation}. We call predicate-based cut
isolation, which prevents Phantoms, \textit{Predicate Cut
  Isolation}.\footnote{Oracle provides an isolation level called
  ``Statement Level Read Consistency''; this is analogous to Predicate
  Cut Isolation at the level of a single operation within a
  multi-operation transaction. We do not consider this isolation model
  here as it is subsumed by standard Snapshot Isolation and we believe
  our discussion is easily extended to incorporate it.}

@@

Perhaps perplexingly, discussions of TA are absent from existing
discussions of weak isolation. This is perhaps again due to the single-node
context in which prior work was developed: on a single server (or a
fully replicated database), TA is near-trivial to achieve via
lightweight locking and/or local concurrency control over data
items. In contrast, in a distributed environment, TA over arbitrary
groups of non-colocated items is considerably more difficult to
achieve with high availability: servers need to know when all of a
transaction's updates are present on their respective
replicas. However, replicas do not need to agree on when to reveal a
new value to clients, which would require consensus; accordingly, TA
only requires reliable broadcast (with some additional metadata for
clients) and is achievable in a HAT system. We omit a full discussion
of the algorithm and instead refer the reader to the extended version
of this paper~\cite{hat-tr}.

@@

Our models have not yet considered interactions across transactions:
we have not guaranteed any ordering or continuity between transactions
(other than that there exists some arbitrary ordering). However, in
the distributed systems context, many useful \textit{safety}
guarantees span multiple application actions. In particular,
\textit{session guarantees} are used to describe guarantees across
groups of operations (here, transactions) issued by the same logical
user~\cite{sessionguarantees}. A \textit{session} is an application
context that consists of logical state that should persist between
transactions. For example, on a social networking site, all of a
user's transactions submitted between ``log in'' and ``log out''
operations might form a session.

There are several guarantees that we can make with high availability:

\vspace{.5em}\noindent\textit{{Monotonic reads}} requires that,
with a session, subsequent reads to a given object ``never return any
previous values''; reads from each item progress forward in a total
order. The ordering of reads should respect any total ordering on
transactions.

@@

We can achieve the above guarantees by forcing servers to wait to
reveal new writes until their respective dependencies are fulfilled on
all replicas. This mechanism effectively requires that all clients
read from a globally agreed upon lower bound on the versions written
to the system. This is highly available as a client will never block
due to inability to find a server with a sufficiently up-to-date
version of a data item. However, it does not imply that transactions
will read their own writes or, in the presence of partitions, make
forward progress through the version history. The problem is that, if
a server becomes partitioned, under the highly available model, we
must handle the possibility that an unfortunate client will be forced
to issue her next requests against the partitioned server!

@@

Read your writes is unavailable in a highly available system. Consider
a client that executes the following two transactions in succession:
\vspace{-.5em}
\begin{align*}
\small
T_1 &: w_x(1)
\\T_2 &: r_x(a)
\end{align*}
If the client executes $T_1$ against a server that is partitioned from
the rest of the other servers, the server must allow $T_1$ to
commit. If the client executes $T_2$ against the same (partitioned)
server, then it will be able to read its writes. However, if the
network topology shifts and the client can only contact a different
server which is partitioned from the server that executed $T_1$, then
the client will be inable to read its own writes and the system will
have to either stall indefinitely to allow the client to read her
writes (violating transactional availability) or will have to
sacrifice read your writes guarantees. Accordingly, read your writes
requires stickiness, and, because causal consistency and PRAM require
read your writes, they also require stickiness. Read your writes is
``free'' in a sticky system, while the remaining causality and PRAM
guarantees can be accomplished with the prior algorithms.

\subsubsection{Additional Guarantees}

@@

While there are infinitely many HAT models
(Section~\ref{sec:futurework}), at this point, we have largely
exhausted the range of achievable, previously defined (and useful) HAT
semantics. Before summarizing our possibility results, we will present
impossibility results for HATs, also defined in terms of previously
identified isolation and consistency anomalies. Perhaps most notably,
it is impossible to prevent Lost Update or Write Skew in a HAT system.

@@

It is impossible to prevent Lost Update in a highly available
environment. Consider two clients who submit the following $T_1$ and
$T_2$ on opposite sides of a network partition:
\begin{align*}
\small
T_1 &: r_x(100)~w_x(100+20=120)
\\T_2 &: r_x(100)~w_x(100+30=130)
\end{align*}
The final state of the database will converge to an execution that
could not have resulted from a serial excecution.\footnote{In this
  example, we assume that, as is standard in modern databases,
  replicas accept values as they are written (i.e., register
  semantics). This particular example could be made serializable via
  the use of commutative updates (Section~\ref{sec:eval}) but, as we
  show in the extended version of the paper~\cite{hat-tr}, the problem
  persists in the general case.} To prevent this from happening,
  either $T_1$ or $T_2$ should not have committed. Each client's
  respective server might try to detect that another write occurred,
  but this requires knowing the version of the latest write to $x$. In
  our example, this reduces to a requirement for linearizability,
  which is, via Gilbert and Lynch's proof of the CAP Theorem, provably
  unachievable with high availabilty~\cite{gilbert-cap}.

@@

The inability to prevent Lost Update means that Consistent Read,
Snapshot Isolation, and Cursor Stability guarantees are all
unavailable. The inability to prevent Lost Update or Write Skew means
that Repeatable Read and One-Copy Serializability guarantees are
unavailable.

@@

Data storage systems make various recency guarantees on reads of data
items. As we have discussed, one of the most famous is
linearizability~\cite{herlihy-art}, which states that reads will
return the last completed write to a data item, and there are several
other (weaker) variants such as safe and regular register
semantics. When applied to transactional semantics, the combination of
one-copy serializability and linearizability is called \textit{strong
  (or strict) one-copy serializability}~\cite{adya} (e.g., Google's
Spanner~\cite{spanner}). It is also common, particularly in systems
that allow reading from masters and slaves, to provide a guarantee
such as ``read a version that is no more than five seconds out of
date'' or similar. Unfortunately, an indefinitely long partition can
force a system to violate any recency bound, so recency bounds are
unavailable to HAT systems.

@@

To the best of our knowledge, this is the first unification of
database ACID, distributed consistency, and session guarantee
models. Perhaps satisfactorily, strong one-copy serializability
subsumes all other models, while considering the (large) power set of
all compatible models (e.g., the diagram depicts 96 possible highly available
combinations) hints at the vast expanse of consistency models found in
the literature. This taxonimization is \textit{not} exhaustive, but we
believe it lends substantial clarity into the relationships between a
large subset of the prominent ACID and distributed consistency
models. Additional read/write transaction semantics that we have
omitted should be easily classifiable based on the available
primitives and unavailable anomalies we have already discussed.

@@

Composition cost also varies by property. For instance, Charron-Bost
has proven that, to capture causality between $N$ communicating
processes, standard vector-based approaches face an upper bound of
$O(N)$ storage per write~\cite{charron-bost}. This means that, with
$100K$ clients, each write might be accompanied by $100K$ timestamps
per vector. This is difficult to scale. By compromising on
availability (e.g., treating a datacenter as a linearizable cluster),
this overhead can be reduced~\cite{cops, eiger}, but it is much
cheaper to provide, say, read your writes, than full causal
consistency.

\vspace{.5em}\noindent\textbf{Linearizability and Transactional
  Atomicity} The relationship between linearizability and
transactional atomicity is non-obvious. One might attempt to find a
mapping between the two: transactional atomicity dictates that writes
to multiple keys across multiple servers are made visible to readers
all at once, while linearizability dictates that writes to a single
key on multiple servers are made visible to all readers at once. The
difference is two-fold. First, in linearizable (and safe and regular)
systems, writes are made visible to all clients \textit{immediately}
after they finish. With transactional atomicity, there is no recency
guarantee. Second, in linearizabile systems, all clients see all
writes at the same time. With transactional atomicity, clients see
writes at different times depending on which replicas they contact. We
are not aware of an analogous model in the distributed systems
literature. Accordingly, despite apparent similarities, transactional
atomicity is much weaker than linearizability.

\vspace{.5em}\noindent\textbf{Visibility and Stickiness} Sticky
availability can result in much better write \textit{visibility}:
clients will be able to safely read writes more quickly in a sticky
available system. In the model we discussed, it is possible to achieve
several properties like monotonic reads in a highly available system
by waiting to reveal a write until all servers have seen it and its
relevant dependencies. However, this incurs severe visibility
penalties---new writes will not become visible to clients in the
presence of partitions. A client that does not want to guarantee
read-your-writes (due to the sticky availability requirement) may
still wish to read other clients' writes with timeliness. Accordingly,
a system may forgo high availability in order to provide better
visibility. We also note that a system may wish to provide stickiness
in the absence of partitions, but, in the presence of partitions,
``lose'' session guarantees by having clients reconnect to an
available replica. This latter strategy provides session guarantees in
the absence of partitions but, to provide availability, sacrifices
them in the event of partitions.

>>>>

HAT systems, or systems providing transactions with transactional
availability with high availability ($1$-available) or sticky high
availability ($1$-sticky-available), offer substantial latency and
availability benefits, yet they come with a cost to achievable
semantics. In this section, we delineate which of several ACID,
distributed consistency, and session consistency levels can be
achieved with high availability (transactional atomicity, variants of
Repeatable Read isolation, and many session guarantees) and sticky
high availability (read your writes, PRAM and causal consistency) and
which cannot (preventing Lost Update and Write Skew, recency).  We
present a full summary of these results in
Section~\ref{sec:hat-summary}.

@@

To begin, Adya's \textit{Read Uncommitted} isolation requires a total
order among transactions and requires that each transaction's writes
are ordered consistently with this order (prohibiting ``Dirty
Writes,'' or $G0$)~\cite{adya}. If two transactions write to the same
set of data items, then the final database state cannot contain the
``earlier'' transaction's writes to the data items. For example, in
the below example, $T_3$ should eventually only read $a=b=1$ or
$a=b=2$ but not $a=2, b=1$ or $a=1, b=2$:
\begin{align*}
\small
T_1 &: w_x(1)~w_y(1)
\\T_2 &: w_x(2)~w_y(2)
\\T_3 &: r_x(a)~r_y(b)\\[-2em]
\end{align*}
We will strengthen this property in later models, which will prevent,
for instance, reading $a=2, b=1$ at any time (not just
``eventually''). The difference between this total order on
transactions and the total order required by serializability is that
the Read Uncommitted total order is completely arbitrary (i.e., an
order must exist), whereas a serializable total order must be
equivalent to a serial execution. Read Uncommitted is easily achieved
via applying the same logical timestamp to each update in a
transaction and applying a ``last writer wins'' conflict
reconciliation policy at each replica.

\textit{Read Committed} isolation requires that transactions do not
read uncommitted versions of data items (prohibiting both ``Dirty
Writes''---as above---and ``Dirty Reads'' phenomena; captured by
Adya's $G1a-c$, ANSI's $P1$, and ``broad'' $P1$ (2.2) from Berenson et
al.). For instance, in the example below, $T_3$ should never see
$a=1$, and, if $T_2$ aborts, $T_3$ should never read $a=3$:
\vspace{-.5em}
\begin{align*}
\small
T_1 &: w_x(1)~w_x(2)
\\T_2 &: w_x(3)\\
T_3 &: r_x(a)\\[-2em]
\end{align*}
It is fairly easy to prvent ``Dirty Reads'': if each transaction never
writes uncommitted data to the database, then transactions will never
read each others' dirty data. As a simple solution, clients can buffer
their writes until they commit, or, alternatively, can send them to
servers, who will not serve new writes until clients notify them that
the writes have been committed.

\textit{Repeatable Read} isolation is a contentious property. As
Berenson et al. discuss~\cite{ansicritique}, Gray's original
Repeatable Read lock-based implementation provides substantially
richer guarantees than those that are required by the ANSI SQL
specification~\cite{gray-isolation}. Gray~\cite{gray-isolation},
Berenson et al.~\cite{ansicritique}, and Adya~\cite{adya} all
interpret Repeatable Read as providing serializability for all
operations except predicate-based reads. However, in its description
of Repeatable Read, the ANSI SQL specification provides a useful
property which, although is not true to Gray's spirit of Repeatable
Read, is also commonly found in distributed consistency models: the
ANSI Repeatable Read guarantee requires that, along with observing
Read Committed isolation, each transaction will only read one version
of each data item that it did not itself produce (preventing ``Fuzzy
Read,'' or P2). In the example below, $T_3$ must read $a=1$:
\begin{align*}
\small
T_1 &: w_x(1)
\\T_2 &: w_x(2)
\\T_3 &: r_x(1)~r_x(a)
\end{align*}
This isolation property is a literal interpretation of the phrase
``Repeatable Read'': unless a transaction modifies a given data item,
the observable value of the data item should not change during the
transaction. By itself, this property is rather weak---we can always
read $\bot$ for each data item. However, when coupled with additional
properties, like transactional atomicity, it is stronger: the repeated
reads must obey additional ordering constraints
(Section~\ref{sec:discussion}). In fact, distributed systems often
describe a ``consistent snapshot'' across a set of related events as a
\textit{cut} across a set of participants or, in our case, data
items. Accordingly, to capture the notion that the transaction should
read from a (non-changing) consistent cut over data items (and to
disambiguate from the aforementioned stronger Repeatable Read
properties), we call ANSI SQL ``Repeatable Read'' (i.e., preventing
P2) \textbf{Cut Isolation}. It is possible to satisfy cut isolation
with high availability by caching appropriate versions for each of a
transaction's reads or, alternatively, using multi-versioning on each
replica and ensuring that each of a transaction's successive reads
return the same version of each data item.

We can further consider two variants of Cut Isolation. The first is
relatively straightforward and provides a cut across individual data
items. We will call this \textit{Item Cut Isolation}. The second
provides a cut over individual data items and over logical ranges of
data items, or predicate-based reads (e.g., \texttt{SELECT
  WHERE}). This prevents the \textit{Phantom Problem}, whereby two
successive predicate-based reads return different
data~\cite{gray-isolation}. We call predicate-based cut isolation,
which prevents Phantoms, \textit{Predicate Cut
  Isolation}.\footnote{Oracle provides an isolation level called
  ``Statement Level Read Consistency''; this is analogous to Predicate
  Cut Isolation at the level of a single operation within a
  multi-operation transaction. We do not consider this isolation model
  here as it is subsumed by standard Snapshot Isolation and we believe
  our discussion is easily extended to incorporate it.}

@@

Perhaps perplexingly, discussions of TA are absent from existing
discussions of weak isolation. This is perhaps again due to the
single-node context in which prior work was developed: on a single
server (or a fully replicated database), TA is achievable via
lightweight locking and/or local concurrency control over data
items. In contrast, in a distributed environment, TA over arbitrary
groups of non-colocated items is considerably more difficult to
achieve with high availability: servers need to know when all of a
transaction's updates are present on their respective
replicas. However, replicas do not need to agree on when to reveal a
new value to clients, which would require consensus; accordingly, TA
only requires reliable broadcast (with some additional metadata for
clients) and is achievable in a HAT system. We omit a full discussion
of the algorithm and instead refer the reader to the extended version
of this paper~\cite{hat-tr}.

@@

Our models have not yet considered interactions \textit{across}
transactions: we have not guaranteed any ordering or continuity
between transactions (other than that there exists some arbitrary
ordering). This is a natural concern in a distributed context, and, in
the distributed systems literature, many useful \textit{safety}
guarantees span multiple application actions. In particular,
\textit{session guarantees} are used to describe guarantees across
transactions within a given \textit{session}, ``an abstraction for the
sequence of...operations performed during the execution of an
application''~\cite{sessionguarantees}. Informally a session describes
a context that should persist between transactions: for example, on a
social networking site, all of a user's transactions submitted between
``log in'' and ``log out'' operations might form a session.

There are several session guarantees that we can make with high availability:

\vspace{.5em}\noindent\textit{{Monotonic reads}} requires that,
within a session, subsequent reads to a given object ``never return any
previous values''; reads from each item progress forward in a total
order. The ordering of reads should respect any total ordering on
transactions.

@@

We can achieve the above guarantees by forcing servers to wait to
reveal new writes until each write's respective dependencies are
fulfilled on all replicas. This mechanism effectively ensures that all
clients read from a globally agreed upon lower bound on the versions
written to the system. This \textit{is} highly available as a client
will never block due to inability to find a server with a sufficiently
up-to-date version of a data item. However, it does not imply that
transactions will read their own writes or, in the presence of
partitions, make forward progress through the version history. The
problem is that, if a server becomes partitioned, under the highly
available model, we must handle the possibility that an unfortunate
client will be forced to issue her next requests against the
partitioned server!

@@

Read your writes is not achievable in a highly available
($1$-available) system. Consider a client that executes the following
two transactions in succession:
\vspace{-.5em}
\begin{align*}
\small
T_1 &: w_x(1)
\\T_2 &: r_x(a)
\end{align*}
If the client executes $T_1$ against a server that is partitioned from
the rest of the other servers, the server must allow $T_1$ to
commit. If the client executes $T_2$ against the same (partitioned)
server, then it will be able to read its writes. However, if the
network topology shifts and the client can only contact a different
server which is partitioned from the server that executed $T_1$, then
the client will be unable to read its own writes and the system will
have to either stall indefinitely to allow the client to read her
writes (violating transactional availability) or will have to
sacrifice read your writes guarantees. However, if the client remains
sticky with the server that executed $T_1$, then we can disallow this
scenario. Accordingly, read your writes, and, by proxy, causal
consistency and PRAM require stickiness. Read your writes is ``free''
in a sticky system, while the remaining causality and PRAM guarantees
can be accomplished with the algorithms for achieving the remaining
session guarantees.

\subsubsection{Additional HAT Guarantees}

In this section, we briefly discuss additional noteworthy guarantees
achievable by HAT systems.

@@

While there are infinitely many HAT models
(Section~\ref{sec:futurework}), at this point, we have largely
exhausted the range of achievable, previously defined (and useful)
semantics that are available to HAT systems. Before summarizing our
possibility results, we will present impossibility results for HATs,
also defined in terms of previously identified isolation and
consistency anomalies. Perhaps most notably, it is impossible to
prevent Lost Update or Write Skew in a HAT system.

@@

It is impossible to prevent Lost Update in a highly available
environment. Consider two clients who submit the following $T_1$ and
$T_2$ on opposite sides of a network partition:
\begin{align*}
\small
T_1 &: r_x(100)~w_x(100+20=120)
\\T_2 &: r_x(100)~w_x(100+30=130)
\end{align*}
Regardless of whether $x=120$ or $x=130$ is chosen by a replica, the
database state could not have arisen serial execution of $T_1$ and
$T_2$.\footnote{In this example, we assume that, as is standard in
  modern databases, replicas accept values as they are written (i.e.,
  register semantics). This particular example could be made
  serializable via the use of commutative updates
  (Section~\ref{sec:eval}) but, as we show in the extended version of
  the paper~\cite{hat-tr}, the problem persists in the general case.}
To prevent this from happening, either $T_1$ or $T_2$ should not have
committed. Each client's respective server might try to detect that
another write occurred, but this requires knowing the version of the
latest write to $x$. In our example, this reduces to a requirement for
linearizability, which is, via Gilbert and Lynch's proof of the CAP
Theorem, provably unachievable with high
availabilty~\cite{gilbert-cap}.

@@

The inability to prevent Lost Update means that Consistent Read,
Snapshot Isolation, and Cursor Stability guarantees are all
unavailable. The inability to prevent Lost Update or Write Skew means
that Repeatable Read and One-Copy Serializability guarantees are
unavailable~\cite{adya,ansicritique}.

@@

Data storage systems make various recency guarantees on reads of data
items. As we have discussed, one of the most famous is
linearizability~\cite{herlihy-art}, which states that reads will
return the last completed write to a data item, and there are several
other (weaker) variants such as safe and regular register
semantics. When applied to transactional semantics, the combination of
one-copy serializability and linearizability is called \textit{strong
  (or strict) one-copy serializability}~\cite{adya} (e.g., Google's
Spanner~\cite{spanner}). It is also common, particularly in systems
that allow reading from masters and slaves, to provide a guarantee
such as ``read a version that is no more than five seconds out of
date'' or similar. Unfortunately, an indefinitely long partition can
force a system to violate any recency bound, so recency bounds are not
enforceable in HAT systems.

@@

To the best of our knowledge, this is the first unification of
database ACID, distributed consistency, and session guarantee
models. Interestingly, strong one-copy serializability subsumes all
other models, while considering the (large) power set of all
compatible models (e.g., the diagram depicts 96 possible highly
available combinations) hints at the vast expanse of consistency
models found in the literature. This taxonimization is \textit{not}
exhaustive, but we believe it lends substantial clarity into the
relationships between a large subset of the prominent ACID and
distributed consistency models. Additional read/write transaction
semantics that we have omitted should be easily classifiable based on
the available primitives and HAT-incompatible anomaly prevention we
have already discussed.

@@

Composition cost also varies by combination. For instance, Charron-Bost
has proven that, to capture causality between $N$ communicating
processes, standard vector-based approaches face an upper bound of
$O(N)$ storage per write~\cite{charron-bost}. This means that, with
$100K$ clients, each write might be accompanied by $100K$ timestamps
per vector. This is difficult to scale. By compromising on
availability (e.g., treating a datacenter as a linearizable cluster),
this overhead can be reduced~\cite{cops, eiger}, but it is much
cheaper to provide, say, read your writes, than full causal
consistency.

\vspace{.5em}\noindent\textbf{Linearizability and Transactional
  Atomicity} The relationship between linearizability and
transactional atomicity is non-obvious. Transactional atomicity
dictates that writes to multiple keys across multiple servers are made
visible to readers all at once, while linearizability dictates that
writes to a single key on multiple servers are made visible to all
readers at once---what is different? First, in linearizable (and safe
and regular) systems, writes are made visible to all clients
\textit{immediately} after they finish. With transactional atomicity,
there is no recency guarantee. Second, in linearizabile systems, all
clients see all writes at the same time. With transactional atomicity,
clients may see writes at different times depending on which replicas
they contact. We are not aware of an analogous model in the
distributed systems literature. Accordingly, despite apparent
similarities, transactional atomicity incomparable with and much
cheaper (by availability standards) than linearizability.

\vspace{.5em}\noindent\textbf{Visibility and Stickiness} Sticky
availability can result in much better write \textit{visibility}:
clients will be able to safely read writes more quickly in a sticky
available system. In the model we discussed, it is possible to achieve
several properties like monotonic reads in a highly available system
by waiting to reveal a write until all servers have seen it and its
relevant dependencies. However, this incurs severe visibility
penalties---new writes will not become visible to clients in the
presence of partitions. A client that does not want to guarantee
read-your-writes (due to the sticky availability requirement) may
still wish to read other clients' writes with timeliness.
